---
title: "Methods Paper Complex Model"
author: "Zachary Esses Johnson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

$~$

**Libraries**
```{r, warning = FALSE, message = FALSE}
library(lavaan)
library(MASS)
library(dplyr)
library(ggplot2)
library(tidyr)
```

$~$

**Set Workpath**
```{r}
workpath <- "C:/Users/zejpo/OneDrive/Documents/Work/Methods Paper"

setwd(workpath)
```

$~$

**Set Seed**
```{r}
set.seed(999)
```

$~$

**Simulation Parameters**
```{r}
real_params <- list(
  n_sims = 100, # Iterations
  sample_sizes = c(200, 500, 1000, 2000, 5000), # Sample sizes
  error_levels = c(0.2, 0.4, 0.6),  # Error levels
  factor_distributions = c("Normal", "Skewed"), # Factor distributions
  heteroskedasticity = c("Homoskedastic", "Heteroskedastic") # Error variance
)
```

$~$

**Real Covariance Matrix**
```{r}
real_cov_matrix <- matrix(0.5, ncol = 4, nrow = 4) # Matrix for 4 factors
diag(real_cov_matrix) <- 1 # Set covariance matrix diagonal to 1
```

$~$

**Real Simulation Loop**
```{r, warning = FALSE}
# Dataframe to store results
real_results <- data.frame()

# Runs simulations 100 times
for(sim in 1:real_params$n_sims) {
  
  # Loop for varied sample sizes
  for(n in real_params$sample_sizes) {
    
    # Loop for varied error levels
    for(error in real_params$error_levels) {
      
      # Loop for varied factor distributions
      for(factor_dist in real_params$factor_distributions) {
        
        # Loop for heteroskedasticity
        for(hetero in real_params$heteroskedasticity) {
          
          # Generate matrix of true latent factors
          true_factors <- if (factor_dist == "Skewed") {
            matrix(rgamma(n * 4, shape = 2, scale = 1), ncol = 4)  # Gamma distribution for skewed factors
          } else {
            mvrnorm(n, mu = rep(0, 4), Sigma = real_cov_matrix) # Multivariate normal for normal factors
          }
          
          # Set factor loadings
          loadings <- c(0.6, 0.7, 0.8, 0.9, 0.5, 0.65)
          
          # Generate residual variance
          error_variances <- if (hetero == "Heteroskedastic") c(0.1, 0.3, 0.5, 0.2, 0.4, 0.3) else rep(error, 12)
          
          # Generate error matrix
          error_matrix <- matrix(rnorm(n * 12, sd = sqrt(rep(error_variances, each = n))), nrow = n)
          
          # Generate observed data
          observed_data <- cbind(
            true_factors[,1] * loadings[1] + error_matrix[,1],
            true_factors[,1] * loadings[2] + error_matrix[,2],
            true_factors[,1] * loadings[3] + error_matrix[,3],
            true_factors[,2] * loadings[1] + error_matrix[,4],
            true_factors[,2] * loadings[2] + error_matrix[,5],
            true_factors[,2] * loadings[3] + error_matrix[,6],
            true_factors[,3] * loadings[1] + error_matrix[,7],
            true_factors[,3] * loadings[2] + error_matrix[,8],
            true_factors[,3] * loadings[3] + error_matrix[,9],
            true_factors[,4] * loadings[1] + error_matrix[,10],
            true_factors[,4] * loadings[2] + error_matrix[,11],
            true_factors[,4] * loadings[3] + error_matrix[,12]
          )
          
          # Standardize observed variables and convert to dataframe
          real_data <- as.data.frame(scale(observed_data))
          colnames(real_data) <- paste0("V", 1:12)
          
          # Step 1 CFA model
          real_model_1 <- "
            f1 =~ V1 + V2 + V3
            f2 =~ V4 + V5 + V6
            f3 =~ V7 + V8 + V9
            f4 =~ V10 + V11 + V12
            
            V2 ~~ V3
            V4 ~~ V5
            V7 ~~ V8
            V10 ~~ V12
          "
          
          # Step 2 CFA model
          real_model_2 <- "
            F =~ f1 + f2 + f3 + f4
          "
          
          # Full model merge
          real_model_full <- paste0(real_model_1, "\n", real_model_2)
          
          # Fitting models
          tryCatch({
            
            # Step 1 CFA estimation
            real_fit1 <- cfa(real_model_1, data = real_data, estimator = "MLR")
          
            if(lavInspect(real_fit1, "converged")) {
              
              # Predict factor scores from step 1
              step1_scores <- lavPredict(real_fit1, method = "Regression")
              step1_scores <- scale(step1_scores)
              
              # Step 2 CFA estimation using predicted scores from step 1
              real_fit2 <- cfa(real_model_2, data = as.data.frame(step1_scores), estimator = "MLR")
              
              if(lavInspect(real_fit2, "converged")) {
                
                # Predict second-order factor scores
                step2_scores <- lavPredict(real_fit2, method = "Regression")
                
                # Correlation between predicted and true mean factor scores
                step2_cor <- cor(step2_scores[, 1], rowMeans(true_factors))
                
                # Estimate full hierarchical CFA model
                real_fit_full <- cfa(real_model_full, data = real_data, estimator = "MLR")
                
                if(lavInspect(real_fit_full, "converged")) {
                  
                  # Predict full model factor scores
                  std_scores <- lavPredict(real_fit_full, method = "Regression")
                  
                  # Correlation between predicted and true mean factor scores
                  std_cor <- cor(std_scores[, 1], rowMeans(true_factors))
                  
                  # Calculate RMSE for both methods
                  step2_rmse <- sqrt(mean((scale(step2_scores[, 1]) - scale(rowMeans(true_factors)))^2, na.rm = TRUE))
                  std_rmse <- sqrt(mean((scale(std_scores[, 1]) - scale(rowMeans(true_factors)))^2, na.rm = TRUE))
                  
                  # Bind results to dataframe
                  real_results <- rbind(real_results, data.frame(
                    sim = sim, n = n, error = error, factor_dist = factor_dist, hetero = hetero,
                    step2_cor = step2_cor, std_cor = std_cor,
                    cor_diff = step2_cor - std_cor,
                    step2_rmse = step2_rmse, std_rmse = std_rmse
                  ))
                }
              }
            }
          }, error = function(e) { cat("Error:", e$message, "\n") })
        }
      }
    }
  }
  # Print progress
  cat("Completed simulation", sim, "of", real_params$n_sims, "\n")
}
```

$~$

**Data Save**
```{r}
saveRDS(real_results, "complex_results_real.rds")
```

$~$

**Read Data**
```{r}
real_results <- readRDS("complex_results_real.rds")
```

$~$

**Summary**
```{r}
rmse_real_summary <- real_results %>%
  group_by(n, factor_dist, hetero) %>%
  summarise(
    mean_step2_rmse = mean(step2_rmse, na.rm = TRUE),  
    mean_std_rmse = mean(std_rmse, na.rm = TRUE),  
    .groups = "drop"
  )
print(rmse_real_summary)
```

$~$

**T-Test Between Sequential and Standard RMSE Values**
```{r}
t_test <- t.test(real_results$step2_rmse, real_results$std_rmse, paired = TRUE)
print(t_test)
```
$~$

**Boxplot for RMSE Distributions**
```{r}
# Reshape data
long_results <- real_results %>%
  select(Sequential = step2_rmse, Traditional = std_rmse) %>%
  pivot_longer(cols = everything(), names_to = "Method", values_to = "RMSE")

# Boxplot
ggplot(long_results, aes(x = Method, y = RMSE, fill = Method)) +
  geom_boxplot(width = 0.6, outlier.shape = 16, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Sequential" = "darkgreen", "Traditional" = "red")) +
  labs(
    x = NULL,
    y = "Root Mean Squared Error (RMSE)",
    fill = "Method"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(face = "bold")
  )
```

$~$

**Table for RMSEs across Conditions**
```{r}
latex_table <- knitr::kable(
  rmse_real_summary,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  caption = "Mean RMSE by Sample Size, Factor Distribution, and Heteroskedasticity"
)

cat(latex_table)
```

$~$

**RMSE Density Comparison**
```{r}
ggplot(real_results) +
  geom_density(aes(x = step2_rmse, fill = "Sequential"), alpha = 0.5) +
  geom_density(aes(x = std_rmse, fill = "Traditional"), alpha = 0.5) +
  facet_grid(factor_dist ~ hetero) +
  theme_minimal() +
  labs(
    x = "RMSE",
    y = "Density",
    fill = "Method"
  ) +
  scale_fill_manual(values = c("Sequential" = "darkgreen", "Traditional" = "red"))
```

$~$

**T-Test Between Sequential and Standard Correlation Values**
```{r}
cor_ttest <- t.test(real_results$step2_cor, real_results$std_cor, paired = TRUE)
print(cor_ttest)
```

$~$

**Correlation Differences Across Conditions**
```{r}
ggplot(real_results, aes(x = factor(n), y = cor_diff, fill = factor_dist)) +
  geom_boxplot() +
  facet_wrap(~hetero) +
  theme_minimal() +
  labs(
    x = "Sample Size",
    y = "Correlation Difference (Sequential - Traditional)",
    fill = "Factor Distribution"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




