---
title: "Sequential CFA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequential CFA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

**Sequential Confirmatory Factor Analysis (Sequential CFA)** is a hierarchical
factor analysis approach that estimates each level of the hierarchy
*independently* rather than simultaneously. The key benefits are:

- The full observed-data sample size is preserved at every stage.
- Cross-level covariance structures do not need to be parameterised together.
- Valid estimates can be obtained in cases where traditional or Bayesian CFA
  fails to converge.

This vignette walks through the complete two-stage workflow using simulated
data: simulate → Stage 1 → propagation diagnostics → Stage 2 → correction →
composite index scores.

```{r load}
library(lavaan)
library(scfa)
```

---

## Step 1 — Simulate a Two-Level Hierarchy

We generate data from a known population model with two sub-factors
(`subfactor1`, `subfactor2`) that load onto a higher-level index factor.

```{r simulate}
set.seed(42)

pop_model <- "
  subfactor1 =~ 0.7*x1 + 0.7*x2 + 0.7*x3
  subfactor2 =~ 0.8*x4 + 0.8*x5 + 0.8*x6
  subfactor1 ~~ 1*subfactor1
  subfactor2 ~~ 1*subfactor2
  subfactor1 ~~ 0*subfactor2
  x1 ~~ 0.51*x1; x2 ~~ 0.51*x2; x3 ~~ 0.51*x3
  x4 ~~ 0.36*x4; x5 ~~ 0.36*x5; x6 ~~ 0.36*x6
"

dat <- simulateData(pop_model, sample.nobs = 400, model.type = "sem")
head(dat)
```

---

## Step 2 — Fit the Stage-1 (Sub-factor) CFA

Fit a standard CFA to the observed indicators at the lowest level of the
hierarchy.

```{r stage1}
model_stage1 <- "
  subfactor1 =~ x1 + x2 + x3
  subfactor2 =~ x4 + x5 + x6
"

fit_stage1 <- cfa(model_stage1, data = dat, std.lv = TRUE)
summary(fit_stage1, fit.measures = TRUE)
```

---

## Step 3 — Error-Propagation Diagnostics

Before running Stage 2, assess how much estimation error will propagate
forward. The key quantity is the **factor score reliability** \(\hat\rho_k\):
a value close to 1 means the Stage-1 factor score is nearly error-free; a
value below the threshold (default 0.70) flags a weak link.

```{r diagnostics}
diag_tbl <- scfa_propagation_diagnostics(fit_stage1, threshold = 0.70)
print(diag_tbl)
```

The table reports:

| Column | Meaning |
|--------|---------|
| `I_k` | Fisher information — how much the indicators tell us about the factor |
| `psi_nu` | Propagation variance \(1/I_k\) — noise injected into Stage-2 residuals |
| `phi_k` | Estimated factor variance |
| `rho_k` | Score reliability \(\phi_k I_k / (\phi_k I_k + 1)\) |
| `flag` | `TRUE` if \(\rho_k\) is below the threshold |

You can also visualise the reliabilities:

```{r plot-diag, fig.width = 6, fig.height = 4}
plot(diag_tbl)
```

---

## Step 4 — Extract Stage-1 Factor Scores

Use **Bartlett** scores (asymptotically unbiased) as inputs to Stage 2.

```{r scores}
scores_stage1 <- as.data.frame(lavPredict(fit_stage1, method = "bartlett"))
head(scores_stage1)
```

---

## Step 5 — Fit the Stage-2 (Index) CFA

Use the Stage-1 factor scores as *observed* indicators for the higher-level
index factor.

```{r stage2}
model_stage2 <- "index =~ subfactor1 + subfactor2"

fit_stage2 <- cfa(model_stage2, data = scores_stage1, std.lv = TRUE)
summary(fit_stage2, fit.measures = TRUE)
```

---

## Step 6 — Correct for Propagated Error (Bartlett Scores)

Under Bartlett scores the Stage-2 **loadings** are asymptotically unbiased, but
each Stage-2 **residual variance** is inflated by the Stage-1 propagation
variance \(\hat\psi_{\nu,k} = 1/\hat I_k\). Use `scfa_correct_residuals()` to
recover the true residual variance.

```{r correct-residuals}
adj_theta <- scfa_correct_residuals(fit_stage2, fit_stage1)
adj_theta
```

> If you had used **regression** scores instead, the residual inflation is
> absorbed into an attenuation bias on the loadings. Use
> `scfa_correct_loadings()` in that case.

---

## Step 7 — Extract Final Index Scores

```{r index}
index_scores <- lavPredict(fit_stage2)
head(index_scores)
```

These are the composite index scores for each observation.

---

## Streamlined: Using `run_scfa()`

All the steps above can be wrapped in a single call to `run_scfa()`:

```{r run-scfa}
result <- run_scfa(
  stage_models = list(
    "subfactor1 =~ x1 + x2 + x3
     subfactor2 =~ x4 + x5 + x6",
    "index =~ subfactor1 + subfactor2"
  ),
  data      = dat,
  method    = "bartlett",
  threshold = 0.70,
  cfa_args  = list(std.lv = TRUE)
)

print(result)

# Stage-1 diagnostics
print(result$diagnostics$stage_1)

# Final index scores
head(result$index_scores)

# Residual variance correction (Bartlett workflow)
result$correction
```

The `result` object also contains the full propagation chain:

```{r chain}
result$propagation_chain
```

---

## Three-Level Hierarchy Example

For a three-level hierarchy (item → sub-factor → factor → index), simply pass
three model strings. `scfa_propagate_chain()` accumulates propagation variances
across all stages:

```{r three-level, eval = FALSE}
result3 <- run_scfa(
  stage_models = list(
    # Stage 1: items -> sub-factors
    "sf1 =~ i1 + i2 + i3
     sf2 =~ i4 + i5 + i6",
    # Stage 2: sub-factors -> factors
    "f1 =~ sf1 + sf2
     f2 =~ sf3 + sf4",
    # Stage 3: factors -> index
    "index =~ f1 + f2"
  ),
  data = my_data
)

# Cumulative propagation variance entering Stage 3
result3$propagation_chain[["stage_3"]]
```

---

## References

Essesjohnson, Z. (2024). *Sequential Confirmatory Factor Analysis*. OSF
Preprints. <https://osf.io/preprints/osf/akxtv_v2>
